Biogeography of crop progenitors and wild plant resources in the Late Pleistocene and Early Holocene of West Asia, 21--8.3 ka

### 2025-01-05 analysis/paper.qmd
Add results

---
title: Biogeography of crop progenitors and wild plant resources in Late Pleistocene–Early Holocene West Asia, 21–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first comprehensive reconstructions of the 
  palaeodistributions of 56 plant species found regularly in association with
  early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using hindcast data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on average [X%] [larger|smaller] in 
  the Early Holocene compared to present conditions, indicating [...]. 
  The modelled ranges predict the observed occurrence of these species on 
  archaeological sites with [low|medium|high] accuracy. The regional ubiquity of 
  species in the archaeological record is [not] correlated with the predicted 
  size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Finish modelling
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [ ] Results
    - [ ] Model assessment
    - [ ] Hindcasting
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
  - [ ] Conclusion
- [ ] Figures
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
  - [ ] tbl-climate-periods
- [ ] Clean up package code
- [ ] Appendix with all hindcast predictions
- [ ] References & copyediting
- [ ] Final proofread

@@

climate_periods <- tibble(
  period = c("Last Glacial Maximum", "Heinrich Stadial 1", "Bølling-Allerød",
             "Younger Dryas", "Early Holocene", "Current"),
  code = c("lgm", "hs1", "ba", "yds", "eh", "cur"),
  label = fct_inorder(c("LGM (21 ka)", "HS1 (17.0-14.7 ka)", "BA (14.7-12.9 ka)", 
            "YDS (12.9-11.7 ka)", "EH (11.7-8.3 ka)", "current")),
  start_bp = c(21000, 17000, 14700, 12900, 11700, NA),
  end_bp = c(17000, 14700, 12900, 11700, 8300, NA)
)
```

@@

@@

```{r fig-region}
#| fig-cap: Study region
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  layer_spatial(w_asia, fill = NA) +
  coord_sf(crs = w_asia_albers)
```

@@

@@

# Resample predictor data to common projection and resolution
climate_cube <- st_warp(climate, crs = w_asia_albers, 
                        cellsize = c(pred_res, pred_res))
terrain_cube <- st_warp(terrain_cube, climate_cube)
soil_cube <- st_warp(soil_cube, climate_cube)

@@

```{r extract-predictor-data}
# TODO: factor out
bind_extract <- function(x, y) {
  extract <- st_extract(y, st_transform(x, st_crs(y)))
  cbind(x, select(tibble(extract), -geometry))
}

@@

```{r niche-model}
# Random Forest Niche Model
# With hyperparameters adapted from the example of a down-sampled RF model given 
# by https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
niche_model <- rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger", 
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
    probability = TRUE
  )

@@

<!-- TODO: rewrite to remove thresholding -->
The output of the model is probabilistic.
However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to an estimate of habitat suitability.
To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
We select the threshold value for each model individually, using MaxSSS [as recommended by @LiuEtAl2013].
This also makes it possible to analyses the predictions together as an assemblage.

@@

@@

```{r fig-model-accuracy-vs-n_occ}
#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
flora |>
  mutate(niche_test_fit = map(niche, collect_metrics)) |>
  unnest(niche_test_fit) |>
  filter(.metric == "accuracy") |> # TODO: or roc_auc?
  mutate(label = if_else(.estimate < 0.75, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic") +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
  theme_half_open() 
```

* Modelled ecological niches on current data

## Hindcasting

Sensitivity to climate fluctuations?

@@

* Comparison to archaeological occurrences

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in the Late Pleistocene and Early Holocene of West Asia, 21–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first comprehensive reconstructions of the 
  palaeodistributions of 56 plant species found regularly in association with
  early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using hindcast data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on average [X%] [larger|smaller] in 
  the Early Holocene compared to present conditions, but did [not] change as
  drastically between the Last Glacial Maximum and the Early Holocene. 
  However, the modelled ranges do not predict the observed occurrence of these 
  species on archaeological sites with any degree of accuracy. The regional 
  ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Finish modelling
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [ ] Results
    - [ ] Model assessment
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
  - [ ] Conclusion
- [ ] Figures
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
  - [ ] tbl-climate-periods
- [ ] Clean up package code
- [ ] Appendix with all hindcast predictions
- [ ] References & copyediting
- [ ] Final proofread

@@

# TODO: drop LGM and HS1 since there are no sites 
# (can then maybe increase prediction resolution?)
climate_periods <- tibble(
  period = c("Last Glacial Maximum", "Heinrich Stadial 1", "Bølling-Allerød",
             "Younger Dryas", "Early Holocene", "Current"),
  code = c("lgm", "hs1", "ba", "yds", "eh", "cur"),
  label = fct_inorder(c("LGM (21 ka)", "HS1 (17.0-14.7 ka)", "BA (14.7-12.9 ka)", 
            "YDS (12.9-11.7 ka)", "EH (11.7-8.3 ka)", "current")),
  start_bp = c(21000, 17000, 14700, 12900, 11700, NA),
  end_bp = c(17000, 14700, 12900, 11700, 8300, NA)
)
```

@@

ne_rivers <- read_sf(raw_data("ne"), "ne_10m_rivers_lake_centerlines_scale_rank")

@@

@@

```{r data-archaeo-sites}
archaeo_sites <- summarise(
  archaeo_flora, 
  latitude = first(latitude), 
  longitude = first(longitude), 
  .by = c(site_name, period)
) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = latlong, remove = FALSE)
```

```{r fig-region}
#| fig-cap: Study region (red) and archaeological sites used to generate modelled flora. Solid circles indicate Neolithic sites.
# TODO: better pallette
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "blue") +
  annotation_spatial(ne_lakes, colour = "blue", fill = "lightblue") +
  layer_spatial(w_asia, colour = "red", fill = NA) +
  layer_spatial(archaeo_sites, shape = 1) +
  layer_spatial(filter(archaeo_sites, period %in% neolithic), shape = 16) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers)
```

@@

# Resample predictor data to common projection and resolution
climate_cube <- st_warp(climate, crs = w_asia_albers, 
                        cellsize = c(pred_res, pred_res), segments = 1000)
terrain_cube <- st_warp(terrain_cube, climate_cube, segments = 1000)
soil_cube <- st_warp(soil_cube, climate_cube, segments = 1000)

@@

```{r data-flora-predictors}
# TODO: factor out
bind_extract <- function(x, y) {
  extract <- st_extract(y, st_transform(x, st_crs(y)))
  cbind(x, select(tibble(extract), -geometry))
}

@@

```{r niche-model}
# Random Forest Niche Model
# With hyperparameters adapted from the example of a down-sampled RF model given 
# by https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
niche_model <- rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger", 
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
    # probability = TRUE
  )

@@

<!-- TODO: prediction, hindcasting and hindcasting assessment methodology -->

@@

We trained Random Forest models for `r nrow(flora)` taxa using contemporary occurrence data from GBIF, a random sample of background points, and the predictor variables described in @tbl-predictors.
Substituting the "current" climate predictors for those derived from palaeoclimatic simulations [@BrownEtAl2018], we could then generate hindcast predictions for reconstructed past environments in `r english(nrow(climate_periods))` key climate periods – a total of `r nrow(flora) * nrow(climate_periods)` modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the appendix and accompanying material.

@@

```{r data-model-assessment}
flora <- mutate(
  flora,
  niche_test_metrics = map(niche, collect_metrics)
)

niche_test_metrics <- flora |>
  unnest(niche_test_metrics) |>
  summarise(
    min = min(.estimate),
    max = max(.estimate),
    mean = mean(.estimate),
    sd = sd(.estimate),
    .by = c(.metric)
  )
```

```{r fig-model-accuracy-vs-n_occ}
#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
flora |>
  unnest(niche_test_metrics) |>
  filter(.metric == "accuracy") |>
  mutate(label = if_else(.estimate < 0.80, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic") +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
  theme_minimal_grid() 
```

We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition.
Model accuracy (i.e. proportion of correctly classified test samples) ranged between `r percent(filter(niche_test_metrics, .metric == "accuracy")$min)` and `r percent(filter(niche_test_metrics, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "accuracy")$mean)`.
The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_metrics, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_metrics, .metric == "accuracy")$sd, 3)`
Model accuracy is loosely correlated with the number of occurrences available for training (@fig-model-accuracy-vs-n_occ), though the worst-performing models also included some species with relatively large sample sizes: <!-- TODO e.g. -->.
Test metrics and ROC curves for the individual models are included in the appendix.

```{r niche-archaeo}
# TODO: factor out
# TODO: what about negatives?
archaeo_flora_predictions <- archaeo_flora |>
  filter(taxon %in% flora$taxon) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = latlong) |>
  mutate(
    prediction = map(taxon, \(x) filter(flora, taxon == x)$prediction[[1]]),
    prediction = map2(prediction, geometry, function(pred, geom) {
      sf_point <- st_transform(st_sf(st_sfc(geom, crs = latlong)), w_asia_albers)
      st_extract(pred, sf_point)
    }),
    prediction = map2(prediction, climate_period, function(pred, per) {
      pred |>
        as_tibble() |>
        filter(as.character(period) == per) |>
        transmute(
          p_absent = prediction.FALSE.,
          p_present = prediction.TRUE.,
          pred_present = p_present > p_absent
        )
    })
  ) |>
  unnest(c(prediction))

p_archaeo_pred_correct <- sum(archaeo_flora_predictions$pred_present, na.rm = TRUE) / length(!is.na(archaeo_flora_predictions$pred_present))

archaeo_flora_pred_summary <- archaeo_flora_predictions |>
  as_tibble() |>
  summarise(
    n_present = sum(!is.na(pred_present)),
    n_predicted = sum(pred_present, na.rm = TRUE),
    p_predicted = n_predicted / n_present,
    .by = c(climate_period, site_name, phase_code, age_mid)
  )
```

The ability of the hindcast models to predict the occurrence of specific species at archaeological sites is significantly worse, with only `r percent(p_archaeo_pred_correct)` of presences in archaeobotanical assemblages successfully predicted.
<!-- TODO: negative -->

----

@@

### 2025-01-05 analysis/paper.tex
Add results

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources in Late Pleistocene--Early Holocene West Asia, 21--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources in Late
Pleistocene--Early Holocene West Asia, 21--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first comprehensive reconstructions of the
palaeodistributions of 56 plant species found regularly in association
with early agricultural archaeological sites in West Asia, including the
progenitors of the first crops. We used machine learning to train an
ecological niche model of each species based on its present-day
distribution in relation to climate and environmental variables.
Predictions of the potential ranges of these species at key stages of
the Pleistocene--Holocene transition could then be derived from these
models using hindcast data from palaeoclimate simulations. Species
ranges are predicted to have been on average {[}X\%{]}
{[}larger\textbar smaller{]} in the Early Holocene compared to present
conditions, indicating {[}\ldots{]}. The modelled ranges predict the
observed occurrence of these species on archaeological sites with
{[}low\textbar medium\textbar high{]} accuracy. The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

\begin{verbatim}
Warning: Removed 51 rows containing missing values or values outside the scale range
(`geom_text_repel()`).
\end{verbatim}

@@

\begin{itemize}
\tightlist
\item
  Modelled ecological niches on current data
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Comparison to archaeological occurrences
\end{itemize}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources in the Late Pleistocene and Early Holocene of West Asia, 21--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources in the
Late Pleistocene and Early Holocene of West Asia, 21--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first comprehensive reconstructions of the
palaeodistributions of 56 plant species found regularly in association
with early agricultural archaeological sites in West Asia, including the
progenitors of the first crops. We used machine learning to train an
ecological niche model of each species based on its present-day
distribution in relation to climate and environmental variables.
Predictions of the potential ranges of these species at key stages of
the Pleistocene--Holocene transition could then be derived from these
models using hindcast data from palaeoclimate simulations. Species
ranges are predicted to have been on average {[}X\%{]}
{[}larger\textbar smaller{]} in the Early Holocene compared to present
conditions, but did {[}not{]} change as drastically between the Last
Glacial Maximum and the Early Holocene. The modelled ranges predict the
observed occurrence of these species on archaeological sites with
{[}low\textbar medium\textbar high{]} accuracy. The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

\begin{verbatim}
Warning: Removed 55 rows containing missing values or values outside the scale range
(`geom_text_repel()`).
\end{verbatim}

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition. Model
accuracy (i.e.~proportion of correctly classified test samples) ranged
between 73\% and 98\%, with an average of 91\%. The area under the
models' receiver operating characteristic curves (ROC-AUC) was in the
range of 0.9734227±0.0596337 Model accuracy is loosely correlated with
the number of occurrences available for training
(Figure~\ref{fig-model-accuracy-vs-n_occ}), though the worst-performing
models also included some species with relatively large sample sizes: .
Test metrics and ROC curves for the individual models are included in
the appendix.

Overall we assessed all models as performing adequately enough to make
reliable predictions on the desired scale, and that in general the
majority of the models had high predictive power. Their predictive
capability when applied to the past is a seperate question, which we
address in the next section.

@@

\begin{itemize}
\tightlist
\item
  Ability to predict archaeological occurrences
\end{itemize}
